An Open-Source Implementation of the CMPS Algorithm for Assessing  Similarity of Bullet Signatures

### 2021-05-28 CMPSpaper_writing/CMPSpaper_writing.Rmd
revised the algorithm section: added the description of the functions into the algorithm section

    + **Correlation peaks:** The parameter `npeaks.set` in `extract_feature_cmps(...)` determines the number of candidate positions and which strategy to  use to combine candidates: If `npeaks.set` is an integer vector of length 1, for example `npeaks.set = 5`, the positions of the top five peaks in the cross-correlation curve are identified as candidates for registration. This strategy is referred as "multi-peak inspection" in @cmps.

    +  **Multiple-length segments:** If `npeaks.set` is an integer vector of length more than 1, for example `npeaks.set = c(5, 3, 1)`, a multi segment lengths strategy will be used: finding top five peaks in the original scale, double the segment length and recalculate the cross-correlation function, then identify 3 peaks in the second scale and repeat the process for a basis segment that is doubled again to identify 1 peak in the third scale. \autoref{fig:seg_all} shows an example of 3 scales of basis segment 6 and their corresponding cross-correlation curves and identified peaks. If a single position (with a tolerance zone determined by `Tx`) results in a top peak in the cross-correlation curve in all three scales, it is called a consistent correlation peak and identified as a candidate position for registration. Note that in @cmps the last element of `npeaks.set` in the multi segment lengths strategy is always set to be 1. Additionally, this strategy is also referred as "multi-peak inspection at different segment lengths" in @cmps.

3. **determine the congruent registration position:** \wj{A candidate position "receives" `votes` from basis segments that identify it or a close position within a tolerance zone of `Tx` as a candidate position in step 2. Votes for all candidate positions} are tallied and the position with the highest number of votes gets chosen as the *congruent registration position*. Basis segments with a congruent registration position are called "congruent matching profile segments" (CMPS). The total number of CMPS is the CMPS score of the comparison. \hh{XXX what happens when there are ties after the vote?}

In the CMPS package, these three steps are implemented using a set of key functions:

* `get_seg_scale(segments, nseg, scale = 2)` takes a basis segment and increases its segment length. `segments` is a list containing all basis segments, generated by the function `get_segs(...)`; `nseg` is an integer indicating which basis segment the function should work with. `scale=1` gives the original length of the basis segment. Increasing `scale` by 1 will double the segment length.

* `get_ccr_peaks(comp, segments, seg_scale, nseg = 1, npeaks = 5)` computes the cross-correlation curve between a basis or scaled segment and the comparison signature and finds peaks in the cross-correlation curve. The number of peaks detected is equal to `npeaks`, which is an integer. `segments`, `seg_scale`, and `nseg` determines the segment in the cross-correlation computation, and `comp` gives the comparison signature. If the multi segment lengths strategy is used, then `get_ccr_peaks(...)` is called in a `lapply()` for each scale of the basis segment. The resulting list is called `ccr.list`.

* `get_ccp(ccr.list, Tx = 25)` tries to identify the "consistent correlation peak" (defined in algorithm step 2) under the framework of multi segment lengths strategy. `ccr.list` is described in the previous bullet point, and `Tx` determines the size of a tolerance zone used in identifying the consistent correlation peak. \hh{XXX what happens when there is no consistent correlation peak?} `get_ccp(...)` returns `NULL` if there is no consistent correlation peak.

* `get_CMPS(input.ccp, Tx = 25)` \hh{XXX how is `get\_CMPS` different from the extract function?} tallies votes from all basis segments and determines the congruent registration position. A position "receives" a vote from a basis segment if this basis segment identifies this position or a closed-enough position (within a tolerance zone defined by `Tx`) as a candidate registration position in algorithm step 2. The goal is to find a shift in registration position that is agreed by the most number of basis segments, so the position receiving the most number of votes is the congruent registration position. In case of ties the middle position is taken as the congruent registration position. Basis segments with the congruent registration position identified are called "congruent matching profile segments" (CMPS). The total number of CMPS is the CMPS score.

Note that there are \hh{several} parameters in the CMPS algorithm \hh{that are left to} the users \hh{to decide}, such as the length of basis segments `seg_length` in step 1, the number of peaks `npeaks.set` identified on each scale level in step 2, and the length of the tolerance zone `Tx` in both step 2 and 3. 
In our implementation of the CMPS algorithm we used the parameters given in the original CMPS paper [@cmps] as the default values for these parameters. \hh{XXX However, the authors state that no cross-validation has been done - there might be also issues with respect to resolution of the scans. Further research is needed, until then users are advised to think of default values as starting values and consider alternatives}
<!--Additionally, as shown in @cmps, using multi segment lengths strategy reduces the number of "false positive" peaks in the cross-correlation curves, thus providing a more conservative way of identifying candidate positions.-->
The main function that combines all steps in the CMPS algorithm described above is called `extract_feature_cmps(...)`. Default values and some explanations of its parameters are as following:

@@

* `x` and `y` are the two bullet signatures. Usually we take `x` as the reference signature (that will be divided into basis segments) and `y` as the comparison signature;
* `seg_length` specifies the length of basis segments [(in mm?)] mentioned in algorithm step 1; 
* `Tx` determines the size of the tolerance zone mentioned in both step 2 and 3;
* `npeaks.set` is a (vector of) integer value(s). This parameter determines which strategy to use to generate candidate positions and the number of peaks to be identified;
* `include` determines the output of the function. Besides the `CMPS.score`, many details of the comparison might be used to understand how the `CMPS.score` is computed. By default, `include = NULL` and the function output will only contain the `CMPS.score`; `include` can also be (an abbreviation of) one of or a vector of the following strings: `"nseg", "congruent.pos", "congruent.seg.idx", "segments", "parameters"`, and `"full_result"`. if `"full_result"` (or its abbreviation) is specified, the ouput will include everything listed below.
  - `nseg`: the number of basis segments we get from the reference signature; this is also the highest possible CMPS score of the comparison;
  - `congruent.pos`: the congruent registration position;
  - `congruent.seg.idx`: the indices of all congruent matching profile segments;
  - `ccp.list`: a list showing identified candidate positions of all basis segments; 
  - `pos.df`: a data frame containing all candidate positions and the number of votes they received;
  - `segments`: a list containing all basis segments;
  - `parameters`: a list containing all input arguments of `extract_feature_cmps`;

@@

<!-- ```{r, echo=FALSE, out.width="400px", fig.cap="put segment\\_plot and scale\\_ccf\\_plot of all scales together. We are identifying five highest peaks at scale 1, three peaks at scale 2, and one peak at scale 3 since npeaks.set = c(5, 3, 1). Putting these plots together can help better understand why segment 6 is not a congruent matching profile segment. Here we can also see clearly that increasing segment length can reduce the number of false positive peaks."} -->
<!-- knitr::include_graphics("img/segplot3.png", dpi = 100) -->
<!-- ``` -->

@@

<!-- Here are the CMPS scores of all 36 signature comparisons of the bullets in the example above. -->

<!-- ```{r} -->
<!-- #      [,1] [,2] [,3] [,4] [,5] [,6] -->
<!-- # [1,]    2    3    2    3    1    2 -->
<!-- # [2,]    2    1   17    1    2    3 -->
<!-- # [3,]    1    1    3   14    2    2 -->
<!-- # [4,]    2    1    1    1   10    1 -->
<!-- # [5,]    2    2    1    1    1   15 -->
<!-- # [6,]   16    3    1    2    1    1 -->
<!-- ``` -->

>>>>

    + **Correlation peaks:** Two strategies referred to as "multi-peak inspection" and "multi-peak inspection at different segment lengths" in @cmps are used for identifying positions of correlation peaks as candidate positions. The latter is also called "multi segment lengths strategy". The parameter `npeaks.set` in `extract_feature_cmps(...)` determines which strategy to use and the number of candidate positions: 
        -   If `npeaks.set` is an integer vector of length 1, for example `npeaks.set = 5`, the positions of the top five peaks in the cross-correlation curve are identified as candidate positions for registration. 
        -   If `npeaks.set` is an integer vector of length more than 1, for example `npeaks.set = c(5, 3, 1)`, the multi segment lengths strategy will be used: finding top five peaks in the original scale, double the segment length and recalculate the cross-correlation function, then identify 3 peaks in the second scale and repeat the process for a basis segment that is doubled again to identify 1 peak in the third scale. \autoref{fig:seg_all} shows an example of 3 scales of basis segment 6 and their corresponding cross-correlation curves and identified peaks.
        -   `get_ccr_peaks(comp, segments, seg_scale, nseg = 1, npeaks = 5)` computes the cross-correlation curve between a basis or scaled segment and the comparison signature and finds peaks in the cross-correlation curve. The number of peaks detected is equal to `npeaks`, which is an integer. `segments`, `seg_scale`, and `nseg` determines the segment in the cross-correlation computation, and `comp` gives the comparison signature. If the multi segment lengths strategy is used, then `get_ccr_peaks(...)` is called in a `lapply()` for each scale of the basis segment. The resulting list is called `ccr.list`.
    +  **Multi segment lengths strategy:** with the multi segment lengths strategy being used, a position is identified as a candidate position for registration and is called as a "consistent correlation peak" if it results in a top peak in the cross-correlation curve with a tolerance zone determined by `Tx` in all segment scales. Note that in @cmps, a segment at its largest scale always identifies one peak.
        -   the scaled version of a basis segment is obtained by using function `get_seg_scale(segments, nseg, scale = 2)`. `segments` is a list containing all basis segments, generated by the function `get_segs(...)` in step 1; `nseg` is an integer indicating which basis segment the function should work with. `scale=1` gives the original length of the basis segment. Increasing `scale` by 1 will double the segment length. 
        -   `get_ccp(ccr.list, Tx = 25)` tries to identify the "consistent correlation peak". `ccr.list` is the result of `lapply()` and `get_ccr_peaks(...)`, and `Tx` determines the size of a tolerance zone used in identifying the consistent correlation peak. `get_ccp(...)` returns `NULL` if there is no consistent correlation peak.

3. **determine the congruent registration position:** A candidate position "receives" `votes` from basis segments that identify it or a close position within a tolerance zone of `Tx` as a candidate position in step 2. Votes for all candidate positions are tallied, and the position with the highest number of votes gets chosen as the *congruent registration position*, \wj{XXX indicating that most of the basis segments find their highly similar counterpart in the comparison signature in terms of correlation at this registration position.} In case of ties the middle position is taken as the congruent registration position. Basis segments with a congruent registration position are called "congruent matching profile segments" (CMPS). The total number of CMPS is the CMPS score of the comparison. `get_CMPS(input.ccp, Tx = 25)` is the function that tallies the votes and determines the congruent registration position and congruent matching profile segments (cmps).

<!-- In the CMPS package, these three steps are implemented using a set of key functions: -->

Note that there are \hh{several} parameters in the CMPS algorithm \hh{that are left to} the users \hh{to decide}, such as the length of basis segments `seg_length` in step 1, the number of peaks `npeaks.set` identified on each scale level in step 2, and the length of the tolerance zone `Tx` in both step 2 and 3. 
In our implementation of the CMPS algorithm we used the parameters given in the original CMPS paper [@cmps] as the default values for these parameters. \hh{XXX However, the authors state that no cross-validation has been done - there might be also issues with respect to resolution of the scans. Further research is needed, until then users are advised to think of default values as starting values and consider alternatives}
<!--Additionally, as shown in @cmps, using multi segment lengths strategy reduces the number of "false positive" peaks in the cross-correlation curves, thus providing a more conservative way of identifying candidate positions.-->
The main function that combines all steps in the CMPS algorithm described above is called `extract_feature_cmps(...)`. Here we present it with its default parameters.

@@

We have discussed parameters `seg_length`, `npeaks.set`, and `Tx` in the previous section, and the parameters left are `x`, `y`, and `include`:

* `x` and `y` are the two bullet signatures. Usually we take `x` as the reference signature (that will be divided into basis segments) and `y` as the comparison signature;

<!-- * `seg_length` specifies the length of basis segments [(in mm?)] mentioned in algorithm step 1;  -->
<!-- * `Tx` determines the size of the tolerance zone mentioned in both step 2 and 3; -->
<!-- * `npeaks.set` is a (vector of) integer value(s). This parameter determines which strategy to use to generate candidate positions and the number of peaks to be identified; -->

* `include` determines the output of the function. Besides the `CMPS.score`, many details of the comparison might be used to understand how the `CMPS.score` is computed. By default, `include = NULL` and the function output will only contain the `CMPS.score`; `include` can also be (an abbreviation of) one of or a vector of the following strings: `"nseg", "congruent.pos", "congruent.seg.idx", "segments", "parameters"`, and `"full_result"`. if `"full_result"` (or its abbreviation) is specified, the ouput will include everything listed below.
  - `nseg`: the number of basis segments we get from the reference signature; this is also the highest possible CMPS score of the comparison;
  - `congruent.pos`: the congruent registration position;
  - `congruent.seg.idx`: the indices of all congruent matching profile segments;
  - `ccp.list`: a list showing identified candidate positions of all basis segments; 
  - `pos.df`: a data frame containing all candidate positions and the number of votes they received;
  - `segments`: a list containing all basis segments;
  - `parameters`: a list containing all input arguments of `extract_feature_cmps`;

@@

@@
